---
title: "Accessibility to COVID-19 vaccination centres in Singapore"
author: "Allan Chong"
editor: visual
execute: 
  warning: false
  message: false
---

## Overview

In a bid to combat Covid-19, Singapore has roll out, Starting from January 4 2023, anyone who needs a Covid-19 immunization or a booster shot can come into any vaccination centers for their shots (The Straits Times, 2023). Residents could visit *Vaccine Go Where* (<https://www.gowhere.gov.sg/vaccine>) to look for their nearest vaccination center based on their postal code, a web application designed by Government Technology Agency (GovTech) of Singapore. (The Straits Times, 2023)

In this exercise, we attempt to look at the Geospatial properties of accessibility of vaccination centers and how residents might prefer one vaccination center over another using Geospatial analytics.

We load the required library for the exercise below:

...

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, sf, spdep, GWmodel, tmap, tidyverse, gtsummary, SpatialAcc, ggstatsplot, reshape2)
```

## Geospatial Data Wrangling

Using the URA Master Plan 2014 subzone boundary GIS data, we will load the spatial data with st_read()

```{r}
mpsz = st_read(dsn="data/geospatial", layer="MP14_SUBZONE_WEB_PL")
```

We then transform the dataset to the SVY21 projection system and ensure that it is valid with the below code

```{r}
mpsz_svy21 = st_transform(mpsz, 3414)
```

We will then need to create a 250m radius hexagons GIS data. This data set was created by using [*st_make_grid()*](https://r-spatial.github.io/sf/reference/st_make_grid.html) of sf package, we will also create an ID for every hexagon.

```{r}
#| eval: false
grid_250  <- st_as_sfc(mpsz_svy21) %>%
  st_make_grid(square = FALSE, cellsize = c(2.5e2, 2.5e2)) %>% 
  st_sf() %>% 
  mutate(id_250 = 1:nrow(.)) # this will be "final" id



```

We save the hexagon file into a new RDS

```{r}
#| eval: false
saveRDS(grid_250, "data/geospatial/hexagon.rds")
```

```{r}
grid_250 = readRDS("data/geospatial/hexagon.rds")
ttm()

tm_shape(grid_250) +
            tmap_options(check.and.fix = TRUE) +
            tm_polygons("id_250", alpha = 0.1) +
            tm_view(set.zoom.limits = c(10,15))
            
#plot(grid_250)
```

```{r}
tm_shape(mpsz_svy21)+
  tm_polygons("PLN_AREA_N", alpha = 0.1) +

tm_shape(mpsz_svy21) +
  tm_polygons("SUBZONE_N", alpha = 0.2) +
  
  tm_view(set.zoom.limits = c(11,15))
```

### Public Transport Points

LTA Datamart provides us with the public transport locations, we will attempt to load them here for analysis

```{r}
busStop = st_read(dsn="data/geospatial/BusStopLocation", layer="BusStop")
busStop = st_transform(busStop, 3414)
```

```{r}
mrt = st_read(dsn="data/geospatial/TrainStationExit", layer="Train_Station_Exit_Layer")
mrt = st_transform(mrt, 3414)
```

```{r}
taxi = st_read(dsn="data/geospatial/TaxiStand", layer="TaxiStop")
taxi = st_transform(taxi, 3414)
```

## Aspatial Data Wrangling

```{r}
vacc_centers = read_csv("data/aspatial/vacc_center_cleaned.csv") 
```

We can use `st_as_sf`to create a dataframe from the longitude (*x*) and latitude (*y*) values. The EPSG 4326 code is used as the dataset is referencing WGS84 geographic coordinate system. We could use `st_crs()`to verify the coordinate system from the object.

```{r}
vacc_centers_sf = st_as_sf(vacc_centers, coords = c("Longitude", "Latitude"), crs=4326)
st_crs(vacc_centers_sf)
```

```{r}
#| eval: false
vacc_centers_sf = st_transform(vacc_centers_sf, 3414)
saveRDS(vacc_centers_sf, "data/aspatial/vacc_centers_sf.rds")
```

```{r}
vacc_centers_sf = readRDS("data/aspatial/vacc_centers_sf.rds")
```

## Reference

The Straits Times (2023), *All can walk in for Covid-19 jabs at vaccination centres from Jan 4* https://www.straitstimes.com/singapore/people-can-walk-into-vaccine-centres-to-get-covid-19-jab-from-wednesday
